---
title: Efficency of Trajectories on Different Spacial Scales
subtitle: Analysis of the GeoLife Trajectory Dataset
author: Manuel Oberhänsli
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
lang: en  # switch to "de" if you write your report in german
bibliography: bibliography.bib
editor: 
  markdown: 
    wrap: 72
---

```{r preprocessing}
#| code-summary: preprocessing
#| eval: false

library(tidyr)
library(readr)
library(sf)
library(tmap)
library(purrr)
library(ggplot2)
library(dplyr)


### READ ALL FILES ###

# get all .plt files
files <- list.files(
  "Data/Raw",
  pattern = "\\.plt$",
  full.names = TRUE
)

# read all files into separate variables
counter <- 1

for(i in seq_along(files)) {
  
  # read trajectory first
  traj <- read_csv(
    files[i],
    skip = 6,
    col_names = c(
      "lat",
      "lon",
      "unused",
      "altitude_ft",
      "days",
      "date",
      "time"
    ),
    show_col_types = FALSE
  ) %>%
    mutate(
      datetime = as.POSIXct(
        paste(date, time),
        format = "%Y-%m-%d %H:%M:%S",
        tz = "UTC"
      )
    )
  
  # skip files with more than 5000 rows
  if(nrow(traj) > 5000) {
    cat("Skipping", files[i], "- too many rows\n")
    next
  }
  
  # create variable name only for kept trajectories
  obj_name <- sprintf("traj_%03d", counter)
  
  # assign into global environment
  assign(obj_name, traj)
  
  counter <- counter + 1
}


### VISUALIZE ALL TRAJECTORIES ###
traj_names <- ls(pattern = "^traj_")

# convert all trajectories to sf LINESTRINGs
traj_lines <- map(
  traj_names,
  function(x) {
    
    traj <- get(x)
    
    # convert to sf points
    traj_sf <- st_as_sf(
      traj,
      coords = c("lon", "lat"),
      crs = 4326
    )
    
    # convert points to line
    traj_line <- traj_sf %>%
      summarise(do_union = FALSE) %>%
      st_cast("LINESTRING")
    
    traj_line$trajectory <- x
    
    traj_line
  }
)

# combine all lines into one sf object
traj_all <- do.call(rbind, traj_lines)

# interactive mode
tmap_mode("view")

# map
tm_shape(traj_all) +
  tm_lines(
    col = "blue",
    lwd = 2
  )


### EXTRACT ACTIVE TRAVEL PARTS ###

timeToNextPoint5min <- function(traj_x) {
  # make sure ordered by time
  traj_x <- traj_x[order(traj_x$datetime), ]
  n <- nrow(traj_x)
  # output vector
  time_to_next_5min <- rep(NA_real_, n)
  for(i in 1:n) {
    current_time <- traj_x$datetime[i]
    # time differences in seconds
    diffs <- as.numeric(
      difftime(
        traj_x$datetime[(i+1):n],
        current_time,
        units = "secs"
      )
    )
    # first point at least 5 min (=300 sec) ahead
    idx <- which(diffs >= 300)[1]
    if(!is.na(idx)) {
      time_to_next_5min[i] <- diffs[idx]
    }
  }
  # add column
  traj_x$timeToNextPoint5min_sec <- time_to_next_5min
  return(traj_x)
}

distToNextPoint5min <- function(traj_x) {
  # make sure ordered by time
  traj_x <- traj_x[order(traj_x$datetime), ]
  n <- nrow(traj_x)
  # output vector
  time_to_next_5min <- rep(NA_real_, n)
  for(i in 1:n) {
    current_time <- traj_x$datetime[i]
    # time differences in seconds
    diffs <- as.numeric(
      difftime(
        traj_x$datetime[(i+1):n],
        current_time,
        units = "secs"
      )
    )
    # first point at least 5 min (=300 sec) ahead
    idx <- which(diffs >= 300)[1]
    if(!is.na(idx)) {
      time_to_next_5min[i] <- diffs[idx]
    }
  }
  # add column
  traj_x$timeToNextPoint5min_sec <- time_to_next_5min
  return(traj_x)
}

distToNextPoint5min <- function(traj_x) {
  # ensure ordering
  traj_x <- traj_x[order(traj_x$datetime), ]
  n <- nrow(traj_x)
  # output vector (meters)
  dist_to_next_5min <- rep(NA_real_, n)
  # convert to sf points (WGS84)
  sf_pts <- st_as_sf(traj_x, coords = c("lon", "lat"), crs = 4326)
  # loop over points
  for(i in 1:n) {
    current_time <- traj_x$datetime[i]
    # find future points
    future_idx <- which(traj_x$datetime > current_time)
    if(length(future_idx) == 0) next
    # time diffs (seconds)
    dt <- as.numeric(difftime(traj_x$datetime[future_idx],
                              current_time,
                              units = "secs"))
    # first index >= 5 minutes
    j <- which(dt >= 300)[1]
    if(!is.na(j)) {
      target_idx <- future_idx[j]
      # distance in meters using sf
      dist <- st_distance(sf_pts[i, ], sf_pts[target_idx, ], by_element = TRUE)
      dist_to_next_5min[i] <- as.numeric(dist)
    }
  }
  traj_x$distToNextPoint5min_m <- dist_to_next_5min
  return(traj_x)
}

speedToNextPoint5min <- function(traj_x){
  traj_x$speedToNextPoint5min_ms <- traj_x$distToNextPoint5min_m / traj_x$timeToNextPoint5min_sec
  traj_x$distToNextPoint5min_m[is.na(traj_x$speedToNextPoint5min_ms)] <- NA
  return(traj_x)
}

# speed threashold: 0.833 m/s (=3km/h)
travelling <- function(traj_x) {
  traj_x$travelling <- traj_x$speedToNextPoint5min_ms >= 0.833
  traj_x$travelling[is.na(traj_x$speedToNextPoint5min_ms)] <- NA
  
  return(traj_x)
}

extractMovingSegment <- function(traj_x) {
  # find all TRUE positions
  moving_idx <- which(traj_x$travelling == TRUE)
  # if no movement exists
  if(length(moving_idx) == 0) {
    return(NULL)
  }
  # first moving point
  start_idx <- moving_idx[1]
  # find first FALSE after movement started
  after_start <- traj_x$travelling[start_idx:nrow(traj_x)]
  end_false <- which(after_start == FALSE)[1]
  # determine end index
  if(is.na(end_false)) {
    end_idx <- nrow(traj_x)
  } else {
    end_idx <- start_idx + end_false - 2
  }
  # return only consecutive moving segment
  traj_moving <- traj_x[start_idx:end_idx, ]
  return(traj_moving)
}

for(nm in traj_names) {
  
  cat("Processing", nm, "...\n")
  
  # load trajectory
  traj <- get(nm)
  
  # processing pipeline
  traj <- timeToNextPoint5min(traj)
  traj <- distToNextPoint5min(traj)
  traj <- speedToNextPoint5min(traj)
  traj <- travelling(traj)
  
  # extract moving segment
  traj_moving <- extractMovingSegment(traj)
  
  # save objects in workspace
  assign(nm, traj)
  assign(paste0(nm, "_moving"), traj_moving)
  
  # export to csv
  write.csv(
    traj_moving,
    file = file.path(
      "Data/Preprocessed",
      paste0(nm, "_moving.csv")
    ),
    row.names = FALSE
  )
}
```

## Code

```{r}
#| echo: true
#| fig.show: hide
#| results: hide

library(tidyr)
library(readr)
library(sf)
library(tmap)
library(purrr)
library(ggplot2)
library(dplyr)

### READ ALL FILES ###

# get all .plt files
files <- list.files(
  "Data/Preprocessed",
  pattern = "\\.csv$",
  full.names = TRUE
)

# read all files into seperate variables
for(i in seq_along(files)) {
  obj_name <- sprintf("traj_%03d", i)
  traj <- read_csv(files[i]) 
  assign(obj_name, traj)
}

names_traj <- ls(pattern = "^traj_")

# delete empty files and files with one point only
empty_trajs <- names_traj[sapply(names_traj, function(x) {
  nrow(get(x)) < 2
})]

rm(list = empty_trajs)
names_traj <- ls(pattern = "^traj_")

# convert all to sf object
for (nm in names_traj) {
  traj <- get(nm)

  traj_sf <- st_as_sf(
    traj,
    coords = c("lon", "lat"),
    crs = 4326,
    remove = FALSE
  )

  assign(nm, traj_sf)
}

all_traj <- bind_rows(
  mget(names_traj),
  .id = "traj_id"
)

# combine all trajectories
all_lines <- all_traj %>%
  arrange(traj_id, datetime) %>%
  group_by(traj_id) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")


### TRAJECTORY LENGTH AND DIRECTNESS ###

add_step_distance <- function(traj) {
  traj <- traj %>%
    arrange(datetime)
  # shift geometry to align each point with the next one
  geom_next <- lead(st_geometry(traj))
  # compute point-to-next-point distance (meters)
  traj$step_dist_m <- as.numeric(
    st_distance(traj, geom_next, by_element = TRUE)
  )
  # last point has no next point
  traj$step_dist_m[is.na(traj$step_dist_m)] <- 0
  traj
}
 
# dist to next point
for (nm in names_traj) {
  traj <- get(nm)
  traj <- add_step_distance(traj)
  assign(nm, traj)
}

# L0: direct path from A to B
add_total_distance <- function(traj) {
  traj <- traj %>%
    arrange(datetime)
  first_pt <- traj[1, ]
  last_pt  <- traj[nrow(traj), ]
  total_dist <- as.numeric(
    st_distance(first_pt, last_pt)
  )
  traj$direct_dist <- total_dist
  traj
}

for (nm in names_traj) {
  traj <- get(nm)
  traj <- add_total_distance(traj)
  assign(nm, traj)
}

# L1: actual path from A to B
add_total_path_distance <- function(traj) {
  traj <- traj %>%
    arrange(datetime)
  traj$total_path_dist_m <- sum(traj$step_dist_m, na.rm = TRUE)
  traj
}

for (nm in names_traj) {
  traj <- get(nm)
  traj <- add_total_path_distance(traj)
  assign(nm, traj)
}

# L1/L0: directness of actual path
for (nm in names_traj) {
  traj <- get(nm)
  traj$path_directness <- traj$total_path_dist_m / traj$direct_dist
  assign(nm, traj)
}


### STATISTICAL ANALYSIS ###

# summarize all trajectories to single data frame
traj_summary <- data.frame(
  traj_id = names_traj,
  total_path_dist_m = NA_real_,
  direct_dist = NA_real_,
  path_directness = NA_real_
)

for (nm in names_traj) {
  traj <- get(nm)
  traj_summary$total_path_dist_m[traj_summary$traj_id == nm] <-
    unique(traj$total_path_dist_m)
  traj_summary$direct_dist[traj_summary$traj_id == nm] <-
    unique(traj$direct_dist)
  traj_summary$path_directness[traj_summary$traj_id == nm] <-
    unique(traj$path_directness)
}

# general
summary(traj_summary)

# plot taken path ~ direct dist
ggplot(traj_summary, aes(x = direct_dist, y = total_path_dist_m)) +
 geom_point(alpha = 0.6) +
 labs(
   x = "direct distance [m]",
   y = "actual length of  path [m]"
 ) +
 theme_minimal()

# plot directness ~ direct dist
ggplot(traj_summary, aes(x = direct_dist, y = path_directness)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "direct distance [m]",
    y = "directness of path = actual length / direct distance"
  ) +
  theme_minimal()

# correlation analysis
ggplot(traj_summary,
       aes(x = direct_dist,
           y = path_directness)) +
  geom_point(alpha = 0.5) +
   geom_smooth(method = "loess", se = TRUE) +
   labs(
     x = "direct distance [m]",
     y = "directness of path"
   ) +theme_minimal()
 
 cor.test(traj_summary$direct_dist, traj_summary$path_directness, method = "spearman")

```

## Abstract

Human mobility is constrained by transportation networks and physical
barriers, leading to travel paths that deviate from the straight-line
distance between origins and destinations. This deviation can be
quantified using trajectory directness, defined as the ratio between
travelled distance and Euclidean distance. This study investigates
whether longer journeys exhibit higher directness using a subsample of
79 trajectories from the GeoLife GPS Trajectories Dataset, which
contains long-term GPS records from 182 users. For each trajectory,
total path length and straight-line distance were computed using spatial
operations in R, and a directness ratio was derived. The relationship
between journey length and directness was evaluated using Spearman rank
correlation. Results show a weak but statistically significant negative
correlation (ρ = -0.297, p = 0.008), indicating that longer journeys
tend to be more direct. However, substantial variability is observed,
particularly in short trajectories. This pattern is likely influenced by
infrastructure effects and preprocessing decisions, including trajectory
segmentation at pauses longer than five minutes, which may
differentially affect transport modes. Overall, the findings suggest
that journey length is associated with increased route efficiency, but
that measurement and mode-related biases play an important role in
observed trajectory structure.

## Introduction

Human mobility is a central aspect of everyday life. Although travelers
often seek efficient routes between origins and destinations, real-world
trajectories rarely follow the shortest possible path. Instead, movement
is constrained by transportation networks, topography, and other
physical barriers, causing travelled distances to exceed the
straight-line distance between start and end points.

The difference between travelled distance and straight-line distance can
be used to quantify the *directness* of a trajectory. Directness
provides a simple measure of how efficiently a journey connects its
origin and destination.

This study investigates the directness of real-world travel trajectories
using the GeoLife GPS Trajectories Dataset, a large collection of GPS
tracks recorded by 182 users over more than five years. For each
trajectory, travelled distance and straight-line distance are compared
to calculate a directness ratio. The relationship between journey length
and directness is then examined. It is hypothesized that longer journeys
exhibit higher directness than shorter journeys.

## Material and Methods

The analysis was based on trajectory data from the GeoLife GPS
Trajectories Dataset (Zheng et al. 2009). To reduce computational
complexity while retaining a diverse set of movement patterns, a random
subsample of approximately 100 trajectories was selected for analysis.

Trajectory Preprocessing. Only active travel segments were included in
the study. Movement was classified as active when the recorded speed
exceeded 3 km/h over the following five consecutive minutes. Trajectory
sequences were required to contain a minimum of two GPS observations to
allow the calculation of movement metrics.

Measures of Trajectory Directness. To quantify the directness of
movement, three measures were calculated for each trajectory using the R
package [sf](https://r-spatial.github.io/sf/?utm_source=chatgpt.com)
(Pebesma, 2023):

1.  **Total trajectory length** – the cumulative distance travelled
    along the recorded trajectory.

2.  **Direct length** – the straight-line (Euclidean) distance between
    the trajectory's start and end points.

3.  **Directness ratio** – calculated as the ratio between direct length
    and total trajectory length:

Directness Ratio = Total Trajectory Length / Direct Length

Values close to 1 indicate highly direct movement, whereas lower values
indicate more circuitous trajectories. Trajectory directness was used as
an indicator of movement efficiency, describing how closely observed
movement paths approximated the shortest possible route between origin
and destination.

Statistical Analysis. The relationship between trajectory length and
directness was investigated using Spearman's rank correlation
coefficient. Spearman's correlation was chosen because it does not
assume normally distributed variables and is robust to non-linear
monotonic relationships.

Visualisation. Data exploration and result presentation were conducted
in R. Scatterplots were created using
[ggplot2](https://ggplot2.tidyverse.org/?utm_source=chatgpt.com)
(Wickham, 2016) to visualize relationships between trajectory
characteristics and directness measures. Spatial representations of
trajectories were produced using
[tmap](https://r-tmap.github.io/tmap/?utm_source=chatgpt.com) (Tennekes,
2024), allowing visual assessment of movement patterns and route
geometry.

The analysis was done in R (R Core Team, 2025). ChatGPT (OpenAI, 2026)
was used as a coding assistant.

## Results

After preprocessing and filtering, 79 trajectories remained for
analysis. Trajectory lengths varied substantially, ranging from 8.6 m to
175.6 km. However, the distribution was strongly right-skewed, with most
trajectories being relatively short (\< 5 km) and only a few
trajectories exceeding several tens of kilometers. Map 1 shows the
trajectories that are left after preprocessing:

```{r}
# visualize
tmap_mode("view")

tm_shape(all_lines) +
  tm_lines(col = "traj_id", lwd = 1, legend.show = FALSE) +
  #tm_basemap("OpenStreetMap", alpha = 0.4) +
  tm_layout(
    legend.show = FALSE,
    frame = TRUE
  )
```

Map 1: All 79 trajectories left after preprocessing.

Trajectory directness ranged from 1.0 to 5.6, with a median value of
1.40 and a mean of 1.62. Most trajectories exhibited directness values
between 1 and 2, indicating that travelled distances were typically only
moderately longer than the corresponding straight-line distances. Highly
indirect trajectories (directness \> 2) were rare and occurred
exclusively among very short trajectories (\< 200 m).

```{r}
#| echo: false
summary(traj_summary)

# plot taken path ~ direct dist
ggplot(traj_summary, aes(x = direct_dist, y = total_path_dist_m)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "direct distance [m]",
    y = "actual length of  path [m]"
  ) +
  theme_minimal()

# plot directness ~ direct dist
ggplot(traj_summary, aes(x = direct_dist, y = path_directness)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "direct distance [m]",
    y = "directness of path = actual length / direct distance"
  ) +
  theme_minimal()
```

A Spearman rank correlation analysis revealed a weak but statistically
significant negative relationship between direct distance and trajectory
directness (ρ = -0.297, p = 0.008). Thus, trajectories with larger
origin–destination distances tended to be more direct than shorter
trajectories. This finding supports the hypothesis that longer journeys
exhibit greater directness.

```{r}
#| echo: false
cor.test(traj_summary$direct_dist, traj_summary$path_directness, method = "spearman")
```

Visual inspection of the scatterplot further confirmed this pattern.
While short trajectories showed a wide range of directness values,
longer trajectories were consistently characterized by lower directness
values and clustered close to the minimum possible value of 1.

## Discussion

The results support the hypothesis that longer journeys are more direct,
with a weak but significant negative correlation between direct distance
and trajectory directness (ρ = -0.297, p = 0.008). However, the
relationship is highly scattered, indicating that journey length alone
is a limited predictor of route efficiency.

A likely explanation is that longer trips provide access to more
efficient, infrastructure-driven routes (e.g., highways or rail
networks), increasing optimization opportunities over larger spatial
scales. In contrast, short-distance movement is more affected by local
constraints and detours, leading to higher variability in directness.

Several methodological factors may also bias the results. Trajectories
were split at pauses longer than five minutes, which may fragment trips
containing layovers (e.g., train transfers or EV charging) and remove
intermediate detours. This likely inflates directness estimates and
reduces observed path length.

This bias is not uniform across transport modes: public transport and
electric vehicles are more likely to include long stops, whereas
walking, cycling, and conventional car travel are less affected. As a
result, apparent differences in directness may partly reflect
segmentation artefacts rather than true movement efficiency.

Overall, while the hypothesis is supported, trajectory directness is
strongly influenced by data preprocessing and transport-mode
composition, limiting direct interpretation of efficiency alone.

## References

R Core Team. (2025). *R: A language and environment for statistical
computing* (Version 4.5.1) \[Computer software\]. R Foundation for
Statistical Computing. <https://www.r-project.org/>

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*.
Springer. <https://ggplot2.tidyverse.org/>

Pebesma, E. (2023). *sf: Simple features for R*.
<https://r-spatial.github.io/sf/>

Tennekes, M. (2024). *tmap: Thematic maps in R*.
<https://r-tmap.github.io/tmap/>

OpenAI. (2026). *ChatGPT (June 8, 2026 version)* \[Large language
model\]. <https://chat.openai.com/>

Zheng, Y., Li, Q., Chen, Y., Xie, X., & Ma, W.-Y. (2009). *Understanding
mobility based on GPS data* (GeoLife project). Microsoft Research Asia.
<https://www.microsoft.com/en-us/research/publication/understanding-mobility-based-on-gps-data/>

### Wordcount

<!-- after installing the wordcountadding, remove the line "#| eval: false" -->

```{r}
# wordcountaddin:word_count("index.qmd")
text <- paste(readLines("index.qmd", warn = FALSE), collapse = " ")
word_count <- length(strsplit(text, "\\s+")[[1]])
word_count
```
